---
title: "U.S. College Scorecard Analysis Report"
author: "Peter, Matt, Maria"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
  pdf_document:
    toc: true
  word_document:
    toc: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 7)

# Load required packages
suppressPackageStartupMessages({
  library(dotenv)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(scales)
})

# Load custom functions
source("R/load_data.R")
source("R/clean_data.R")
source("R/define_groups.R")
source("R/visualizations.R")

# Load environment variables from .env file
load_dot_env()
API_KEY <- Sys.getenv("API_KEY")

# Set parameters
# Note: If main.R has already been run, this report will use the cached data
# from the /data directory, and the API key is not strictly needed.
FIELDS <- c(
  "id", "school.name", "school.state", "school.city", "school.ownership",
  "latest.admissions.sat_scores.average.overall",
  "latest.cost.tuition.in_state", "latest.cost.tuition.out_of_state",
  "latest.cost.avg_net_price.overall",
  "latest.completion.rate_suppressed.overall",
  "latest.earnings.10_yrs_after_entry.median",
  "latest.aid.median_debt.completers.overall"
)
FIELDS_STR <- paste(FIELDS, collapse = ",")

# Execute the data processing pipeline
df_raw <- load_scorecard_data(api_key = API_KEY, fields_str = FIELDS_STR)
df_clean <- clean_scorecard_data(df_raw)
df_final <- add_comparison_groups(df_clean)
```

## Introduction

This project analyzes public data from the U.S. College Scorecard to explore differences between various types of universities in terms of academic performance, cost-effectiveness, and graduate earnings. We focus on the following comparison groups:

-   **Ivy League**
-   **Florida Public**
-   **New College of Florida**
-   **Private Nonprofit (National)**
-   **Public (National)**

## Descriptive Statistics

First, we calculate the average values for several key metrics across each group to get a preliminary overview.

```{r summary-table}
group_means <- df_final %>%
  group_by(group_label) %>%
  summarize(
    `Number of Schools` = n(),
    `Mean SAT` = round(mean(sat_avg, na.rm = TRUE)),
    `Mean Net Price` = dollar(mean(net_price, na.rm = TRUE)),
    `Mean Grad. Rate` = percent(mean(grad_rate, na.rm = TRUE), accuracy = 0.1),
    `Mean 10-Yr Earnings` = dollar(mean(earn10, na.rm = TRUE))
  ) %>%
  arrange(desc(`Mean SAT`))

kable(group_means, caption = "Key Metric Averages by Comparison Group")
```


The table reveals several important patterns across institution types. Ivy League schools lead in SAT scores (1539) and post-graduation earnings ($102,655), but also have the highest net price ($24,262). 

Notably, Florida Public universities demonstrate exceptional value - despite having similar SAT scores to Private Nonprofit schools (1193 vs 1202), their graduates actually earn more ($55,776 vs $53,786), while charging less than half the price. This suggests strong return on investment, possibly driven by Florida's robust job market.

Graduation rates show stark inequality across sectors. Ivy League schools graduate 96.2% of students, while Public National institutions graduate only 44.6% - meaning more than half of students at typical public universities don't complete their degrees. This 51.6 percentage point gap represents a significant equity challenge in higher education.

The "Other" category presents the poorest value proposition, charging nearly as much as Private Nonprofit institutions ($20,228) while producing the lowest earnings outcomes ($33,159). Meanwhile, New College of Florida stands out for affordability at just $4,536 - less than half the cost of other public universities - while still delivering respectable earnings of $48,082.

When examining cost-benefit ratios, Private Nonprofit schools cost 2.1 times more than Public schools but deliver only 19% higher earnings, suggesting the private premium may not justify the additional expense for many students focused on earnings potential.

## Visual Analysis

### SAT Score Distribution

This box plot illustrates the distribution of SAT scores for each group. The median and overall range for the Ivy League are significantly higher than for other groups.

```{r plot-sat}
plot_sat_boxplot(df_final)
```

### Median Net Price Comparison

This bar chart compares the median net price across groups. Public universities in Florida and nationwide appear to provide the most cost-effective options.Students at the New College of Florida receive an unusual amount of scholarships, thus accounting for the low average price. 

```{r plot-price}
plot_net_price_barchart(df_final)
```

### Graduation Rate Comparison

Graduation rate is a key indicator of institutional success. Ivy League schools have the highest mean graduation rate, reflecting strong academic support and student quality.

```{r plot-grad-rate}
plot_grad_rate_barchart(df_final)
```

### Post-Graduation Earnings vs. Net Price

This scatter plot explores the return on investment in education. The x-axis represents the net price, the y-axis shows median earnings 10 years after entry, and the size of each point corresponds to the graduation rate.

```{r plot-scatter, fig.height=8}
plot_earnings_scatterplot(df_final)
```

As seen in the chart, while Ivy League schools (red points) are more expensive, their graduates also tend to have the highest earnings, suggesting a strong return on investment.

### TOP 50 colleges Earnings-to-Net-Price

<!-- ```{r plot-map} -->
<!-- print(p_map) -->
<!-- ``` -->

## Conclusion

Through this modular analysis pipeline, we have clearly compared the performance of different university groups. The analysis reveals strong correlations between school type, cost, and the future success of their students.